\large Differentiable Cosmological Simulation with the Adjoint Method \vspace{0.3em}

### 2024-02-06 docs/papers/adjoint/adjoint.tex
Finalize ApJS adjoint paper

\end{abstract}

\vspace{1em}
\section{Introduction}

@@

The first differentiable cosmological simulations, such as BORG, ELUCID,
and BORG-PM \citep{BORG, ELUCID, BORG-PM}, were developed before the
advent of modern AD systems, and were based on \HL{the analytic derivatives,
which involve first convoluted derivation by hand using the chain rule
\citep[see e.g.,][App.~D]{SeljakEtAl2017} before implementing them in
code.}
Later codes including \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM,
vmad, SeljakEtAl2017, FlowPM} compute gradients using the AD engines,
namely \texttt{vmad} (written by the same authors) and
\texttt{TensorFlow}, respectively.
\HL{The AD frameworks automatically apply the chain rule to the primitive
operations that comprise the whole simulation, relieving the burden of
derivation and implementation of the derivatives.}
Both analytic differentiation and AD backpropagate the gradients through
the whole history, which requires saving the states at all time steps in
memory.
Therefore, they are subject to a trade-off between time and space/mass
resolution, usually sacrificing both.
As a result, they lose accuracy on small scales and in dense regions
where the time resolution is important, e.g., in weak lensing
\citep{MADLens}.

\HL{Alternatively, the adjoint method provides systematic ways of deriving
the gradients of an objective function $\cJ$ under constraints
\citep{Pontryagin1962}, such as those imposed by the $N$-body equations
of motion in a simulated Universe.
It identifies a set of \emph{adjoint variables} $\vlambda$, dual to the
state variables $\vz$ of the model, and carrying the gradient
information of the objective function with respect to the model state
$\p\cJ / \p\vz$.
For time-dependent problems, the adjoint variables evolve
\emph{backward} in time by a set of equations dual to that of the
forward evolution, known as the \emph{adjoint equations}.
For continuous time, the adjoint equations are a set of differential
equations, while in the discrete case they become difference equations
which are practically a systematic way to structure the chain rule or
backpropagation.
Their initial conditions are set by the explicit dependence of the
objective on the simulation state, e.g., $\vlambda_n = \p\cJ / \p\vz_n$
if $\cJ$ is a function of the final state $\vz_n$.
Solving the adjoint equations can help us to propagate the gradient
information via the adjoint variables to the input parameters $\vtheta$,
to compute the objective gradients $\d\cJ / \d\vtheta$.
And we will see later that the propagated and accumulated gradients on
parameters come naturally from multiple origins, each reflecting a
$\vtheta$-dependence at one stage of the modeling in
\autoref{fig:model}.}

The backward adjoint evolutions depend on the states in the forward run,
which we can re-simulate with reverse time integration if the dynamics
is reversible, thereby dramatically reducing the memory cost
\citep{NeuralODE}.
Furthermore, we \HL{derive the discrete adjoint equations dual to the
discrete forward time integration, known as the discretize-then-optimize
approach \citep[e.g.,][]{ANODE}, to ensure gradients propagate backward
along the \emph{same} trajectory as taken by the forward time
integration.
This is in contrast with the optimize-then-discretize approach, that
numerically integrates the continuous adjoint equations, and is prone to
larger deviations between the forward and the backward trajectories due
to different discretizations \citep{LanzieriLanusseEtAl2022}.}
In brief, to compute the gradient we only need to evolve a simulation
forward, and then backward jointly with its dual adjoint equations.
We introduce the adjoint method first for generic time-dependent
problems in both continuous and discrete cases in \autoref{sec:bwd}, and
then present its application on cosmological simulation in
\autoref{sec:adj}.

@@

  \node (ic) [right=of boltz] {IC} edge [from] (boltz);

@@

  \node (ic) {IC};

@@

(``Boltz'') and initial condition generator (``IC'') in
\autoref{sec:ic}; force solver ($F$) in \autoref{sec:force}; time
integration (``Integ''), kick ($K$), and drift ($D$) in
\autoref{sec:integ}; observation (``Obs'' and $O$) and objective
(``Obj'') in \autoref{sec:obsobj}.

@@

Alternatively, the adjoint equation can be derived from the variational
equation, using the facts that $\d \bm{M}^{-1} / \d t = - \bm{M}^{-1} \,
\dot{\bm{M}} \, \bm{M}^{-1}$ \HL{for any invertible matrix $\bm{M}$}, and
$(\p\vz / \p\tilde\vz)^{-1} = \p\tilde\vz / \p\vz$.
Like \eqref{variational} and \eqref{adjoint}
%
\begin{equation}
\frac\d{\d t} \frac{\p\vz}{\p\tilde\vz}
= \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\tilde\vz},
\qquad
\frac\d{\d t} \frac{\p\tilde\vz}{\p\vz}
= - \frac{\p\tilde\vz}{\p\vz} \cdot \frac{\p\vf}{\p\vz}.
\end{equation}
%
As we can see next, the adjoint equation takes a similar form when one
optimizes an objective function of the state.

@@

For the $N$-body particles, the state variable\footnote{State and
adjoint vectors in the adjoint equations include enumeration of
particles, e.g., $\vnabla\vvarphi$ includes the $\vnabla\varphi$ of each
particle.} is $\vz = (\vx, \vp)^\intercal$.
Their adjoint variables help to accumulate the objective gradient while
evolving backward in time by the adjoint equation.
Let's denote them by $\vlambda = (\vxi, \vpi)$.
We can compare each step of \eqref{KDFK} and \eqref{O} to
\eqref{explicit}, and write down its adjoint equation following
\eqref{explicit_adj}.
Taking $D_{i-1}^i$ for example, we can write it explicitly as
%
\begin{equation*}
\begin{pmatrix} \vx_i \\ \vp_{i-\half} \end{pmatrix}
= \begin{pmatrix} \vx_{i-1} \\ \vp_{i-\half} \end{pmatrix}
+ \begin{pmatrix}
    \vp_{i-\half} D(t_{i-\half}, t_{i-1}, t_i) \\ \vzero
  \end{pmatrix},
\end{equation*}
%
in the form of \eqref{explicit}.
By \eqref{explicit_adj}, its adjoint equation is
%
\begin{align*}
\begin{pmatrix} \vxi_{i-\half} & \vpi_{i-1} \end{pmatrix}
&= \begin{pmatrix} \vxi_{i-\half} & \vpi_i \end{pmatrix}
+ \begin{pmatrix} \vxi_{i-\half} & \vpi_i \end{pmatrix}
  \begin{pmatrix}
    \vzero & \vI \, D(t_{i-\half}, t_{i-1}, t_i) \\
    \vzero & \vzero \\
  \end{pmatrix} \\
%
&= \begin{pmatrix} \vxi_{i-\half} & \vpi_i \end{pmatrix}
- \begin{pmatrix}
    \vzero & \vxi_{i-\half} D(t_{i-\half}, t_i, t_{i-1})
  \end{pmatrix},
\end{align*}
%
where we have used the fact that $D(t_c, t_a, t_b) = - D(t_c, t_b,
t_a)$, and left the $\p\cJ / \p\vz_i$ term, \HL{the explicit dependence of
the objective on the intermediate states (from, e.g., observables on the
light cone), to the observation operator $O$ below.}
This also naturally determines the subscripts of $\xi$ and $\pi$.

@@

Finally, during the reverse time integration, we can accumulate the
objective gradient following \eqref{explicit_grad}:
%
\begin{align}
\frac{\d\cJ}{\d\vtheta} =& \frac{\p\cJ}{\p\vtheta}
%
  + \frac{\p\cJ}{\p\hat\vx} \cdot \frac{\p\hat\vx}{\p\vtheta}
  + \frac{\p\cJ}{\p\hat\vp} \cdot \frac{\p\hat\vp}{\p\vtheta}
%
  + \vxi_0 \cdot \frac{\p\vx_0}{\p\vtheta}
  + \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \nonumber\\
%
  &- \sum_{i=1}^n \Bigl[
    \bigl( \vpi_i \cdot\va_i \frac\p{\p\vtheta} + \vzeta_i
      \bigr) K(t_i, t_i, t_{i-\half})
    + \vxi_i \cdot \vp_{i-\half}
      \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta} \nonumber\\
    &\qquad + \bigl( \vpi_{i-1} \cdot \va_{i-1} \frac\p{\p\vtheta}
      + \vzeta_{i-1} \bigr) K(t_{i-1}, t_{i-\half}, t_{i-1})
    \Bigr], \nonumber\\
%
\frac{\d\cJ}{\d\vomega} =& \frac{\p\cJ}{\p\vomega}
%
  + \vxi_0 \cdot \frac{\p\vx_0}{\p\vomega}
  + \vpi_0 \cdot \frac{\p\vp_0}{\p\vomega},
\label{KDFKO_grad}
\end{align}
%
\HL{where the latter backpropagates from fewer sources than the former as
shown in \autoref{fig:model}.}
To implement \eqref{KDFKO_adj}-\eqref{KDFKO_grad} in \pmwd\ with
\texttt{JAX}, we only need to write custom VJP rules for the high-level
$N$-body integration-observation loop, while the derivatives and VJPs of
the remaining parts including regularization/prior, the observation, the
initial conditions, the kick and drift factors, the growth and transfer
functions, etc., can all be conveniently computed by AD.

@@

Model differentiation evolves the adjoint equations backward in time.
To save memory, the trajectory of the model state in the forward run is
not saved, but re-simulated together with the adjoint equations.
Even though in principle the $N$-body systems are reversible, in
practice the reconstructed trajectory can differ from the forward one
due to \HL{the finite numerical precision and exacerbated by the chaotic
dynamics.
Better reversibility means the gradients propagate backward along a
trajectory closer to the forward path, and thus would be more accurate.}
To test this, in \autoref{tab:rev} we compare the RMSDs of the
forward-then-reverse particle displacements and velocities from the LPT
initial conditions, relative to their respective standard deviations,
that are very small at the \emph{initial} time.
As before, we vary the floating-point precision, the mesh size, the
particle mass, and the number of time steps.
The order of factor importance and their effects are the same as in the
reproducibility test.
This is because more nonlinear structures are more difficult to reverse.
One way to improve reversibility is use higher order LPT to initialize
the $N$-body simulations at later times \citep{MichauxEtAl2021}, when
the displacements and velocities are not as small.
We leave this for future development.

\begin{table*}[t]
\centering
\caption{\pmwd\ reversibility on GPU.
Our adjoint method reduces memory cost by reconstructing the forward
evolution with reverse time integration.
We test the numerical reversibility by comparing the displacements and
velocities of particles that have evolved to $a=1$ and then reversed to
$a=1/64$, to those of the LPT initial conditions at $a=1/64$, in RMSDs.
We take their ratios to the respective standard deviations, about $0.1
\, \Mpc / h$ and $0.7 \, \mathrm{km} / \mathrm{s}$.
Their smallness is the main reason that the quoted relative differences
here are \HL{orders of magnitude greater than those in \autoref{tab:rep}
(see \autoref{fig:grads} where the reversibility is few times more
important than the reproducibility in their impacts on the gradients).}
With the same setup as that in \autoref{tab:rep}, we see the same
general trend---left factors are more important than the right ones,
and lower rows are more reversible than the upper ones.
}
\label{tab:rev}
\begin{tabular}{cccccc}
  \toprule
  precision & cell/ptcl & ptcl mass $[10^{10} M_\odot]$ & time steps
    & disp rel diff & vel rel diff \\
  \midrule
  single & 8 & 1 &  63 & $5.2\times10^{-2}$  & $7.1\times10^{-2}$  \\
  single & 8 & 1 & 126 & $2.1\times10^{-2}$  & $3.6\times10^{-2}$  \\
  single & 8 & 8 &  63 & $3.3\times10^{-3}$  & $7.6\times10^{-3}$  \\
  single & 8 & 8 & 126 & $3.7\times10^{-3}$  & $7.0\times10^{-3}$  \\
  single & 1 & 1 &  63 & $1.4\times10^{-3}$  & $2.2\times10^{-3}$  \\
  single & 1 & 1 & 126 & $1.3\times10^{-3}$  & $1.7\times10^{-3}$  \\
  single & 1 & 8 &  63 & $4.3\times10^{-4}$  & $7.3\times10^{-4}$  \\
  single & 1 & 8 & 126 & $4.4\times10^{-4}$  & $6.3\times10^{-4}$  \\
  double & 8 & 1 &  63 & $5.4\times10^{-11}$ & $1.3\times10^{-10}$ \\
  double & 8 & 1 & 126 & $3.5\times10^{-11}$ & $7.0\times10^{-11}$ \\
  double & 8 & 8 &  63 & $5.8\times10^{-12}$ & $1.4\times10^{-11}$ \\
  double & 8 & 8 & 126 & $6.4\times10^{-12}$ & $1.2\times10^{-11}$ \\
  double & 1 & 1 &  63 & $2.2\times10^{-12}$ & $3.4\times10^{-12}$ \\
  double & 1 & 1 & 126 & $2.1\times10^{-12}$ & $2.9\times10^{-12}$ \\
  double & 1 & 8 &  63 & $7.2\times10^{-13}$ & $1.2\times10^{-12}$ \\
  double & 1 & 8 & 126 & $6.9\times10^{-13}$ & $9.4\times10^{-13}$ \\
  \bottomrule
\end{tabular}
\end{table*}

Next, we want to verify that our adjoint method yields the same
gradients as those computed by AD.
As explained in \autoref{sec:adj}, \pmwd\ already utilizes AD on most of
the differentiation tasks.
To get the AD gradients we disable our custom VJP implementations on the
$N$-body time integration, and the scatter and gather operations.
In \autoref{fig:grads}, we compare the adjoint and AD gradients on a
smaller problem, because AD already runs out of memory if we double the
number of time steps or increase the space/mass resolution by
$2^3\times$ from the listed specifications in the caption.
\HL{For better statistics, we repeat both adjoint and AD runs for 64 times,
with the same cosmology and white noise modes, and compare their results
by an asymmetric difference of $X_i - Y_j$, where $1 \leq j < i \leq
64$.
First we set $X$ and $Y$ to the adjoint and AD gradients respectively,
and find they agree very well on the real white noise (so do their
gradients on cosmological parameters not shown here; see
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/adjoint/grads.txt}{\faFile}).
In addition, we can set both $X$ and $Y$ to either adjoint or AD to
check their respective \emph{reproducibility}.
We find both gradients are consistent among different runs of
themselves, with AD being a lot more reproducible without uncertainty
from the reverse time integration but only that from the GPU
reproducibility.
This implies that we can ignore the reproducibility errors
(\autoref{tab:rep}) when the reversibility ones dominate
(\autoref{tab:rev}).
Though this statement should be verified again in the future when we
reduce the reversibility errors using, e.g., 3LPT.}
%NOTE: this behavior has disappeared with new JAX/CUDA/GPU
%Interestingly, the asymmetric difference of AD gradients is not
%symmetric about 0, implying some difference between the early and late
%AD results, i.e., possible breaking of order- or time-invariance by
%\texttt{JAX} AD.

\begin{figure*}[t]
\centering
\hspace{7em}
\includegraphics[width=0.5\linewidth]{grads.pdf}
\\
\vspace{1em}
\includegraphics[height=0.45\linewidth]{grads_cmp.pdf}
\hfill
\includegraphics[height=0.45\linewidth]{grads_diff.pdf}
\caption{Adjoint gradients of a $128\times128$ slice of the real white
noise field $\d\cJ / \d\vomega$ (top panel), in a \pmwd\ simulation of
$128^3$ particles in a $(128\,\Mpc/h)^3$ box, with a $256^3$ mesh, 15
time steps, and single precision.
\HL{We choose a mean squared error (MSE) objective between two realizations
with the same cosmology but different initial modes, on their density
fields on the $256^3$ mesh at $a=1$, and then compute the gradients with
respect to one realization while holding the other fixed.}
We compare the adjoint gradients to those by AD, for which we have
disabled the custom gradient implementation on the scatter, gather, and
$N$-body time stepping operators.
The adjoint and AD gradients agree as expected, with a RMSD of $\approx
4\times10^{-5}$, 3 orders of magnitude smaller than the standard
deviation of the gradients itself, $\approx 0.015$.
It is also comparable to the difference between two different adjoint
gradients, with a RMSD $\approx 5\times10^{-5}$.
Different AD gradients are more consistent, with a tighter RMSD of
$\approx 1\times10^{-5}$ due to the absence of uncertainty from reverse
time integration.
}
\label{fig:grads}
\end{figure*}

Our last test on the adjoint gradients uses them in a toy optimization
problem in \autoref{fig:optim}.
We use the Adam optimizer \citep{Adam} with a learning rate of $0.1$ and
the default values for other hyperparameters.
Holding the cosmological parameters fixed, we can optimize the real
white noise modes to source initial particles to evolve into our target
pattern, in hundreds to thousands of iterations.
\HL{Interestingly, we find that the variance of the modes becomes bigger
than 1 (that of the standard normal white noise) after the optimization,
and the optimized modes show some level of spatial correlation not
present in the white noise fields, suggesting that the optimized initial
conditions are probably no longer Gaussian.}

@@

\vspace{1em}
\appendix

>>>>

\clearpage
\end{abstract}

\section{Introduction}

@@

The first differentiable cosmological simulations, such as BORG, ELUCID,
and BORG-PM \citep{BORG, ELUCID, BORG-PM}, were developed before the
advent of modern AD systems, and were based on the analytic derivatives,
which involve first convoluted derivation by hand using the chain rule
\citep[see e.g.,][App.~D]{SeljakEtAl2017} before implementing them in
code.
Later codes including \texttt{FastPM} and \texttt{FlowPM} \citep{FastPM,
vmad, SeljakEtAl2017, FlowPM} compute gradients using the AD engines,
namely \texttt{vmad} (written by the same authors) and
\texttt{TensorFlow}, respectively.
The AD frameworks automatically apply the chain rule to the primitive
operations that comprise the whole simulation, relieving the burden of
derivation and implementation of the derivatives.
Both analytic differentiation and AD backpropagate the gradients through
the whole history, which requires saving the states at all time steps in
memory.
Therefore, they are subject to a trade-off between time and space/mass
resolution, usually sacrificing both.
As a result, they lose accuracy on small scales and in dense regions
where the time resolution is important, e.g., in weak lensing
\citep{MADLens}.

Alternatively, the adjoint method provides systematic ways of deriving
the gradients of an objective function $\cJ$ under constraints
\citep{Pontryagin1962}, such as those imposed by the $N$-body equations
of motion in a simulated Universe.
It identifies a set of \emph{adjoint variables} $\vlambda$, dual to the
state variables $\vz$ of the model, and carrying the gradient
information of the objective function with respect to the model state
$\p\cJ / \p\vz$.
For time-dependent problems, the adjoint variables evolve
\emph{backward} in time by a set of equations dual to that of the
forward evolution, known as the \emph{adjoint equations}.
For continuous time, the adjoint equations are a set of differential
equations, while in the discrete case they become difference equations
which are practically a systematic way to structure the chain rule or
backpropagation.
Their initial conditions are set by the explicit dependence of the
objective on the simulation state, e.g., $\vlambda_n = \p\cJ / \p\vz_n$
if $\cJ$ is a function of the final state $\vz_n$.
Solving the adjoint equations can help us to propagate the gradient
information via the adjoint variables to the input parameters $\vtheta$,
to compute the objective gradients $\d\cJ / \d\vtheta$.
And we will see later that the propagated and accumulated gradients on
parameters come naturally from multiple origins, each reflecting a
$\vtheta$-dependence at one stage of the modeling in
\autoref{fig:model}.

The backward adjoint evolutions depend on the states in the forward run,
which we can re-simulate with reverse time integration if the dynamics
is reversible, thereby dramatically reducing the memory cost
\citep{NeuralODE}.
Furthermore, we derive the discrete adjoint equations dual to the
discrete forward time integration, known as the discretize-then-optimize
approach \citep[e.g.,][]{ANODE}, to ensure gradients propagate backward
along the \emph{same} trajectory as taken by the forward time
integration.
This is in contrast with the optimize-then-discretize approach, that
numerically integrates the continuous adjoint equations, and is prone to
larger deviations between the forward and the backward trajectories due
to different discretizations \citep{LanzieriLanusseEtAl2022}.
In brief, to compute the gradient we only need to evolve a simulation
forward, and then backward jointly with its dual adjoint equations.
We introduce the adjoint method first for generic time-dependent
problems in both continuous and discrete cases in \autoref{sec:bwd}, and
then present its application on cosmological simulation in
\autoref{sec:adj}.

@@

  \node (ic) [right=of boltz] {LPT} edge [from] (boltz);

@@

  \node (ic) {LPT};

@@

(``Boltz'') and initial condition generator by the Lagrangian
perturbation theory (``LPT'') in \autoref{sec:ic}; force solver ($F$) in
\autoref{sec:force}; time integration (``Integ''), kick ($K$), and drift
($D$) in \autoref{sec:integ}; observation (``Obs'' and $O$) and
objective (``Obj'') in \autoref{sec:obsobj}.

@@

Alternatively, the adjoint equation can be derived from the variational
equation, using the facts that $\d \bm{M}^{-1} / \d t = - \bm{M}^{-1} \,
\dot{\bm{M}} \, \bm{M}^{-1}$ for any invertible matrix $\bm{M}$, and
$(\p\vz / \p\tilde\vz)^{-1} = \p\tilde\vz / \p\vz$.
Like \eqref{variational} and \eqref{adjoint}
%
\begin{equation}
\frac\d{\d t} \frac{\p\vz}{\p\tilde\vz}
= \frac{\p\vf}{\p\vz} \cdot \frac{\p\vz}{\p\tilde\vz},
\qquad
\frac\d{\d t} \frac{\p\tilde\vz}{\p\vz}
= - \frac{\p\tilde\vz}{\p\vz} \cdot \frac{\p\vf}{\p\vz}.
\end{equation}
%
As we can see next, the adjoint equation takes a similar form when one
optimizes an objective function of the state.

@@

For the $N$-body particles, the state variable\footnote{State and
adjoint vectors in the adjoint equations include enumeration of
particles, e.g., $\vnabla\vvarphi$ includes the $\vnabla\varphi$ of each
particle.} is $\vz = (\vx, \vp)^\intercal$.
Their adjoint variables help to accumulate the objective gradient while
evolving backward in time by the adjoint equation.
Let's denote them by $\vlambda = (\vxi, \vpi)$.
We can compare each step of \eqref{KDFK} and \eqref{O} to
\eqref{explicit}, and write down its adjoint equation following
\eqref{explicit_adj}.
Taking $D_{i-1}^i$ for example, we can write it explicitly as
%
\begin{equation*}
\begin{pmatrix} \vx_i \\ \vp_{i-\half} \end{pmatrix}
= \begin{pmatrix} \vx_{i-1} \\ \vp_{i-\half} \end{pmatrix}
+ \begin{pmatrix}
    \vp_{i-\half} D(t_{i-\half}, t_{i-1}, t_i) \\ \vzero
  \end{pmatrix},
\end{equation*}
%
in the form of \eqref{explicit}.
By \eqref{explicit_adj}, its adjoint equation is
%
\begin{align*}
\begin{pmatrix} \vxi_{i-\half} & \vpi_{i-1} \end{pmatrix}
&= \begin{pmatrix} \vxi_{i-\half} & \vpi_i \end{pmatrix}
+ \begin{pmatrix} \vxi_{i-\half} & \vpi_i \end{pmatrix}
  \begin{pmatrix}
    \vzero & \vI \, D(t_{i-\half}, t_{i-1}, t_i) \\
    \vzero & \vzero \\
  \end{pmatrix} \\
%
&= \begin{pmatrix} \vxi_{i-\half} & \vpi_i \end{pmatrix}
- \begin{pmatrix}
    \vzero & \vxi_{i-\half} D(t_{i-\half}, t_i, t_{i-1})
  \end{pmatrix},
\end{align*}
%
where we have used the fact that $D(t_c, t_a, t_b) = - D(t_c, t_b,
t_a)$, and left the $\p\cJ / \p\vz_i$ term, the explicit dependence of
the objective on the intermediate states (from, e.g., observables on the
light cone), to the observation operator $O$ below.
This also naturally determines the subscripts of $\xi$ and $\pi$.

@@

Finally, during the reverse time integration, we can accumulate the
objective gradient following \eqref{explicit_grad}:
%
\begin{align}
\frac{\d\cJ}{\d\vtheta} =& \frac{\p\cJ}{\p\vtheta}
%
  + \frac{\p\cJ}{\p\hat\vx} \cdot \frac{\p\hat\vx}{\p\vtheta}
  + \frac{\p\cJ}{\p\hat\vp} \cdot \frac{\p\hat\vp}{\p\vtheta}
%
  + \vxi_0 \cdot \frac{\p\vx_0}{\p\vtheta}
  + \vpi_0 \cdot \frac{\p\vp_0}{\p\vtheta} \nonumber\\
%
  &- \sum_{i=1}^n \Bigl[
    \bigl( \vpi_i \cdot\va_i \frac\p{\p\vtheta} + \vzeta_i
      \bigr) K(t_i, t_i, t_{i-\half})
    + \vxi_{i-\half} \cdot \vp_{i-\half}
      \frac{\p D(t_{i-\half}, t_i, t_{i-1})}{\p\vtheta} \nonumber\\
    &\qquad + \bigl( \vpi_{i-1} \cdot \va_{i-1} \frac\p{\p\vtheta}
      + \vzeta_{i-1} \bigr) K(t_{i-1}, t_{i-\half}, t_{i-1})
    \Bigr], \nonumber\\
%
\frac{\d\cJ}{\d\vomega} =& \frac{\p\cJ}{\p\vomega}
%
  + \vxi_0 \cdot \frac{\p\vx_0}{\p\vomega}
  + \vpi_0 \cdot \frac{\p\vp_0}{\p\vomega},
\label{KDFKO_grad}
\end{align}
%
where the latter backpropagates from fewer sources than the former as
shown in \autoref{fig:model}.
To implement \eqref{KDFKO_adj}-\eqref{KDFKO_grad} in \pmwd\ with
\texttt{JAX}, we only need to write custom VJP rules for the high-level
$N$-body integration-observation loop, while the derivatives and VJPs of
the remaining parts including regularization/prior, the observation, the
initial conditions, the kick and drift factors, the growth and transfer
functions, etc., can all be conveniently computed by AD.

@@

Model differentiation evolves the adjoint equations backward in time.
To save memory, the trajectory of the model state in the forward run is
not saved, but re-simulated together with the adjoint equations.
Even though in principle the $N$-body systems are reversible, in
practice the reconstructed trajectory can differ from the forward one
due to the finite numerical precision and exacerbated by the chaotic
dynamics.
Better reversibility means the gradients propagate backward along a
trajectory closer to the forward path, and thus would be more accurate.
To test this, in \autoref{tab:rev} we compare the RMSDs of the
forward-then-reverse particle displacements and velocities from the LPT
initial conditions, relative to their respective standard deviations,
that are very small at the \emph{initial} time.
As before, we vary the floating-point precision, the mesh size, the
particle mass, and the number of time steps.
The order of factor importance and their effects are the same as in the
reproducibility test.
This is because more nonlinear structures are more difficult to reverse.
One way to improve reversibility is use higher order LPT to initialize
the $N$-body simulations at later times \citep{MichauxEtAl2021}, when
the displacements and velocities are not as small.
We leave this for future development.

\begin{table*}[t]
\centering
\caption{\pmwd\ reversibility on GPU.
Our adjoint method reduces memory cost by reconstructing the forward
evolution with reverse time integration.
We test the numerical reversibility by comparing the displacements and
velocities of particles that have evolved to $a=1$ and then reversed to
$a=1/64$, to those of the LPT initial conditions at $a=1/64$, in RMSDs.
We take their ratios to the respective standard deviations, about $0.1
\, \Mpc / h$ and $0.7 \, \mathrm{km} / \mathrm{s}$.
Their smallness is the main reason that the quoted relative differences
here are orders of magnitude greater than those in \autoref{tab:rep}
(see \autoref{fig:grads} where the reversibility is few times more
important than the reproducibility in their impacts on the gradients).
With the same setup as that in \autoref{tab:rep}, we see the same
general trend---left factors are more important than the right ones,
and lower rows are more reversible than the upper ones.
}
\label{tab:rev}
\begin{tabular}{cccccc}
  \toprule
  precision & cell/ptcl & ptcl mass $[10^{10} M_\odot]$ & time steps
    & disp rel diff & vel rel diff \\
  \midrule
  single & 8 & 1 &  63 & $5.2\times10^{-2}$  & $7.1\times10^{-2}$  \\
  single & 8 & 1 & 126 & $2.1\times10^{-2}$  & $3.6\times10^{-2}$  \\
  single & 8 & 8 &  63 & $3.3\times10^{-3}$  & $7.6\times10^{-3}$  \\
  single & 8 & 8 & 126 & $3.7\times10^{-3}$  & $7.0\times10^{-3}$  \\
  single & 1 & 1 &  63 & $1.4\times10^{-3}$  & $2.2\times10^{-3}$  \\
  single & 1 & 1 & 126 & $1.3\times10^{-3}$  & $1.7\times10^{-3}$  \\
  single & 1 & 8 &  63 & $4.3\times10^{-4}$  & $7.3\times10^{-4}$  \\
  single & 1 & 8 & 126 & $4.4\times10^{-4}$  & $6.3\times10^{-4}$  \\
  double & 8 & 1 &  63 & $5.4\times10^{-11}$ & $1.3\times10^{-10}$ \\
  double & 8 & 1 & 126 & $3.5\times10^{-11}$ & $7.0\times10^{-11}$ \\
  double & 8 & 8 &  63 & $5.8\times10^{-12}$ & $1.4\times10^{-11}$ \\
  double & 8 & 8 & 126 & $6.4\times10^{-12}$ & $1.2\times10^{-11}$ \\
  double & 1 & 1 &  63 & $2.2\times10^{-12}$ & $3.4\times10^{-12}$ \\
  double & 1 & 1 & 126 & $2.1\times10^{-12}$ & $2.9\times10^{-12}$ \\
  double & 1 & 8 &  63 & $7.2\times10^{-13}$ & $1.2\times10^{-12}$ \\
  double & 1 & 8 & 126 & $6.9\times10^{-13}$ & $9.4\times10^{-13}$ \\
  \bottomrule
\end{tabular}
\end{table*}

Next, we want to verify that our adjoint method yields the same
gradients as those computed by AD.
As explained in \autoref{sec:adj}, \pmwd\ already utilizes AD on most of
the differentiation tasks.
To get the AD gradients we disable our custom VJP implementations on the
$N$-body time integration, and the scatter and gather operations.
In \autoref{fig:grads}, we compare the adjoint and AD gradients on a
smaller problem, because AD already runs out of memory if we double the
number of time steps or increase the space/mass resolution by
$2^3\times$ from the listed specifications in the caption.
For better statistics, we repeat both adjoint and AD runs for 64 times,
with the same cosmology and white noise modes, and compare their results
by an asymmetric difference of $X_i - Y_j$, where $1 \leq j < i \leq
64$.
First we set $X$ and $Y$ to the adjoint and AD gradients respectively,
and find they agree very well on the real white noise (so do their
gradients on cosmological parameters not shown here; see
\href{https://github.com/eelregit/pmwd/tree/master/docs/papers/adjoint/grads.txt}{\faFile}).
In addition, we can set both $X$ and $Y$ to either adjoint or AD to
check their respective \emph{reproducibility}.
We find both gradients are consistent among different runs of
themselves, with AD being a lot more reproducible without uncertainty
from the reverse time integration but only that from the GPU
reproducibility.
This implies that we can ignore the reproducibility errors
(\autoref{tab:rep}) when the reversibility ones dominate
(\autoref{tab:rev}).
Though this statement should be verified again in the future when we
reduce the reversibility errors using, e.g., 3LPT.
%NOTE: this behavior has disappeared with new JAX/CUDA/GPU
%Interestingly, the asymmetric difference of AD gradients is not
%symmetric about 0, implying some difference between the early and late
%AD results, i.e., possible breaking of order- or time-invariance by
%\texttt{JAX} AD.

\begin{figure*}[t]
\centering
\hspace{7em}
\includegraphics[width=0.5\linewidth]{grads.pdf}
\\
\vspace{1em}
\includegraphics[height=0.45\linewidth]{grads_cmp.pdf}
\hfill
\includegraphics[height=0.45\linewidth]{grads_diff.pdf}
\caption{Adjoint gradients of a $128\times128$ slice of the real white
noise field $\d\cJ / \d\vomega$ (top panel), in a \pmwd\ simulation of
$128^3$ particles in a $(128\,\Mpc/h)^3$ box, with a $256^3$ mesh, 15
time steps, and single precision.
We choose a mean squared error (MSE) objective between two realizations
with the same cosmology but different initial modes, on their density
fields on the $256^3$ mesh at $a=1$, and then compute the gradients with
respect to one realization while holding the other fixed.
We compare the adjoint gradients to those by AD, for which we have
disabled the custom gradient implementation on the scatter, gather, and
$N$-body time stepping operators.
The adjoint and AD gradients agree as expected, with a RMSD of $\approx
4\times10^{-5}$, 3 orders of magnitude smaller than the standard
deviation of the gradients itself, $\approx 0.015$.
It is also comparable to the difference between two different adjoint
gradients, with a RMSD $\approx 5\times10^{-5}$.
Different AD gradients are more consistent, with a tighter RMSD of
$\approx 1\times10^{-5}$ due to the absence of uncertainty from reverse
time integration.
}
\label{fig:grads}
\end{figure*}

Our last test on the adjoint gradients uses them in a toy optimization
problem in \autoref{fig:optim}.
We use the Adam optimizer \citep{Adam} with a learning rate of $0.1$ and
the default values for other hyperparameters.
Holding the cosmological parameters fixed, we can optimize the real
white noise modes to source initial particles to evolve into our target
pattern, in hundreds to thousands of iterations.
Interestingly, we find that the variance of the modes becomes bigger
than 1 (that of the standard normal white noise) after the optimization,
and the optimized modes show some level of spatial correlation not
present in the white noise fields, suggesting that the optimized initial
conditions are probably no longer Gaussian.

@@

\clearpage
\appendix
